### 2023-07-12 calculus.tex
Updates from Overleaf

We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of the Hazel language server. In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching, strings, and explicit polymorphism (all following the same recipe we developed above). Hazel is implemented in OCaml and compiles to Javascript via \li{js_of_ocaml} \cite{jsoo} for use in the browser. The supplemental material contains a pre-built implementation.

>>>>

We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of the Hazel. % TODO citation and explain what in the world Hazel is (reviewer A stated that the lack of this made this section pretty much impossible to understand)
In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching, strings, and explicit polymorphism (all following the same recipe we developed above).
Hazel is implemented in OCaml and compiles to Javascript via \li{js_of_ocaml} \cite{jsoo} for use in the browser.
The supplemental material contains a pre-built implementation.

### 2023-07-12 intro.tex
Updates from Overleaf

\begin{center}
  \includegraphics[width=0.8\textwidth]{images/haz3l-inconsistent-branches-cursor-inspector.png}
\end{center}

This localization decision affects how the system recovers as it proceeds into the let body. 
If localization had assumed that the \li{then} branch is correct, as for example in \li{merlin}, then 
\li{x : Bool} and an error should be reported on only the second use of \li{x}. 
If the \li{else} branch was chosen, then \li{x : Int} and an error would be reported on only the first use of \li{x}. 
 In either case, this error might mislead the programmer if the earlier localization guess was incorrect.
If the inconsistency is localized to the conditional expression as a whole, as in Hazel, then we again confront the problem of missing type information: 
\li{x} does not have a known type,
though it \emph{is} known to be bound. 
There is no definitive type or binding error at either use of \li{x}, 
so we do not report an error in Hazel. More specifically, we treat \li{x}'s type as the unknown a.k.a. dynamic type from gradual typing.
In any of these cases, we would like to be able to recover and 
localize the type inconsistency on the string addend because the \li{+} operator is integer addition in Hazel.
Recovering from this error, we can assume that both branch types in the second conditional will have \li{Int} type, so no 
error is marked on the conditional as a whole.

@@

Section~\ref{sec:calculus} of this paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus}.
We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
type annotations to expressions \cite{Localinf,BidirTyping}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{Siek06a,siek2015refined}, and 
(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{HazelnutPOPL}, operationalizing the ``red lines'' that an editor displays to the user.
The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solar2013program}) can be marked, i.e. its errors can be simultaneously localized, such that the resulting program has a well-defined type and binding structure.
We establish this and other properties as metatheorems that we mechanize in the Agda proof assistant. 
As we define the calculus, we consider a number of situations where error localization decisions are subtle and conclude by defining some extensions of the intentionally minimal core calculus, e.g. with System-F-style polymorphism and pattern matching, 
that serve as case studies of the more general approach that we hope will be adopted by language designers defining practical type systems.

>>>>

\begin{center}
  \includegraphics[width=0.9\textwidth]{images/haz3l-inconsistent-branches-cursor-inspector.png}
\end{center}

This localization decision affects how the system recovers as it proceeds into the let body. 
If localization had assumed that the \li{then} branch were correct, as for example in \li{merlin}, then 
\li{x : Bool} and an error should be reported on only the second use of \li{x}. 
If the \li{else} branch were chosen, then \li{x : Int} and an error would be reported on only the first use of \li{x}. 
 In either case, this error might mislead the programmer if the earlier localization guess was incorrect.
If the inconsistency were localized to the conditional expression as a whole, as in Hazel, then we again confront the problem of missing type information: 
\li{x} does not have a known type,
though it \emph{is} known to be bound. 
There is no definitive type or binding error at either use of \li{x}, 
so we do not report an error in Hazel. More specifically, we treat \li{x}'s type as the unknown a.k.a. dynamic type from gradual typing.
In any of these cases, we would like to be able to recover and 
localize the type inconsistency on the string addend because the \li{+} operator is integer addition in Hazel.
Recovering from this error, we can assume that both branch types in the second conditional will have \li{Int} type, so no 
error is marked on the conditional as a whole.

@@

Section~\ref{sec:calculus} of this paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus}.
We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
type annotations to expressions \cite{Localinf,BidirTyping}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{Siek06a,siek2015refined}, and 
(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{HazelnutPOPL}, operationalizing the ``red lines'' that an editor displays to the user.
The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solar2013program}) can be marked, i.e. its errors can be simultaneously localized, such that the resulting program has a well-defined type and binding structure.
We establish this and other properties as metatheorems that we mechanize in the Agda proof assistant. 
As we define the calculus, we consider a number of situations where error localization decisions are subtle and conclude by defining some extensions of the intentionally minimal core calculus, e.g. with destructuring patterns and System F-style polymorphism, 
that serve as case studies of the more general approach that we hope will be adopted by language designers defining practical type systems.
